Crunching Large Graphs with Commodity Processors

### 2011-01-26 doc/hotpar2011/softxmt-hotpar2011.tex
a few tweaks

%\documentclass{acm_proc_article-sp}
\documentclass[10pt,nocopyrightspace,preprint]{sigplanconf}
\usepackage{url}
\usepackage{fancyvrb}
\usepackage{alltt}
\usepackage{graphicx}
\usepackage{comment}
\usepackage[labelfont=bf,font=small,belowskip=-7pt,aboveskip=-10pt]{caption}
\begin{document}

%\textheight 9in
%\textwidth 6.5in
%\columnsep 1pc
\renewcommand{\baselinestretch}{0.9}

@@

%todo format the UW names horizontally
\authorinfo{Jacob Nelson, Brandon Myers, A.H. Hunter, Preston Briggs,
  Dan Grossman, Mark Oskin, Carl Ebeling, Simon Kahan, Luis
  Ceze}{University of Washington}{\{nelson, bdmyers, ahh, preston,
  djg, oskin, ebeling, skahan, luisceze\}@cs.washington.edu}

@@

Multithreading is a technique that has been used successfully to
implement efficient computations for these graphs. The Cray XMT is an
example of such an approach: it solves the memory latency problem
through concurrency rather than caching. Each XMT processor supports
128 hardware contexts and 1024 outstanding memory operations, and is
able to switch contexts every cycle \todo{cite XMT description}. This ability comes at a cost: the
XMT is an expensive, non-commodity machine. \todo{don't we want to also say XMT has poor performance on x kind of apps?}

\begin{figure*}[htbp]
  \begin{center}
    \vspace{-0.25in}
    \includegraphics[width=0.7\textwidth]{figures/system-overview.pdf}
    \vspace{-0.2in}
	\end{center}
	\caption{Overview of our proposed system design. The base
          system is a cluster of commodity nodes and interconnect. We
          add a latency-tolerant runtime and a global memory manager
          to obtain efficient access to the global address space. Only
          the shaded components may require hardware design.}
	\label{fig:system-overview}
\end{figure*}

@@

\begin {comment}
\todo{Change to discuss API:

context switches:
  yield()
  discuss coroutines

mem:
  alloc()
  read(p) 
  write(p, d)
  prefetch(p)

sync:
 {read, write} {EE, EF, FE, FF}
}

While the ideal programming model is one that facilitates easy
expression in the problem domain as well as efficient execution on the
available hardware, our programming model today is motivated by only
efficient execution: it exists as a rudimentary library for concept
demonstration, not as a language suitable for development.  The
programming model we use today is therefore only partially formed. It
exists as an augmentation of existing threaded programming models such
as pthreads or openmp that are readily available on most computer
systems.

Though not strictly necessary, for ease of exposition we imagine a
one-to-one binding between threads and cores.  Our model manifests
within any thread independently of the others as a ``fray'' \todo{for lack
of better term }.  Within the thread, a fray instance is the
instantiation of a set of coroutines and a scheduler.  Instantiation
is at user-level and the operating system sees the entire fray as as a
single user-level thread.

\todo{ describe calls that perform thread \& scheduler instantiation; and schedule activation }

Each coroutine executes non-preemptively \todo{seems like non-premptively isn't ``how" it executes, rather how it is scheduled, maybe say ``cooperatively"?}, utilizing the core to which
its ``parent'' thread has been assigned by the operating system until it
reaches a yield point.  Typically, the coroutine yields after
initiating a prefetch operation on data that the programmer or
compiler anticipates would probably not complete before the data is
referenced by a subsequent load operation.  In this way, the core
continues to execute instructions for other coroutines when otherwise
it would likely stall.  The programming model thus tolerates latency.

\todo{ describe calls to prefetch-and-yield }

When a thread frays, for the sake of expediency, a fixed portion of
its stack is divided amongst the coroutines.  Given a finite stack,
each coroutine is constrained to execute a call tree of depth known to
be finite at compilation time: recursion is not yet permitted.

\todo{ discussion of coroutines \& synchronization } Were a coroutine to
block on a synchronization variable shared with other threads, the
entire fray would suspend execution.  This can lead to deadlock when,
for example, one coroutine waits to consume from another thread that
is waiting to consume what only another coroutine in this first fray
can produce.  Instead, coroutines must yield on failed synchronization
events, spinning rather than blocking, where were they bona fide
threads, blocking might be more efficient.

\todo{ say something about ``synchronization'' within a fray }

Coroutines completing their work yield without adding themselves to
the scheduling queue.  The last coroutine to exit in this way returns
as the main thread, as in the common  fork-join model of parallelism.

\end{comment}

\section{Implementation}
\label{sec:approach}
Our goal is to use multithreading to tolerate
the latency of random accesses to memory in a global address space
spread across multiple nodes.  Our implementation must provide three features:
\begin{itemize}
\item {\bf Lightweight multithreading}

  We expect to need hundreds of
  threads to tolerated a cluster's network latency, so our threading
  implementation must have low overhead.

\item {\bf Concurrency in global memory references}

  A thread making a long-latency global memory reference must be able
  to yield and allow other threads to make progress.

\item {\bf Low overhead synchronization}

  We must support global synchronization between threads that allows unrelated threads to proceed unblocked.
\end{itemize}
We discuss our approach to solving each of these challenges in
turn. For each one, we discuss both our current single-node
implementation and ideas for a multi-node implementation.

@@

\subsubsection{Single-node implementation}
We use prefetch instructions for the issue operation, and regular
blocking reads and writes for the complete operation. Since we expect
these global references to have low locality, we use the
``non-temporal'' form of prefetch, which tries to bring the data near
the core without disturbing the state of the cache
hierarchy \cite{intel:swdev}.

@@

We believe we will need hundreds of outstanding memory references per processor to
cover the latency of remote references in a multi-node
system\todo{ can we make a stronger statement? we know it will be 300$+$ according to our calculations}. Unfortunately, commodity processors have much smaller limits
on memory concurrency: section~\ref{subsubsec:evalsinglebase} finds a limit of
36 concurrent loads. We will have to manage memory concurrency on our
own to avoid this limit. We describe approaches in both software and
hardware.

@@

This approach has the benefit of enabling our entire system to be
build using commodity parts, but its potential for performance is not
clear. We have increased the overhead of a single memory operation:
what in the single node case was a simple read, is now multiple
queuing operations between threads and multiple reads and writes over
the processor's link to the network interface.

Another approach moves the global memory management to a hardware
device. We can add a coprocessor FPGA in a processor socket, listening
to coherence messages on the bus. Note that this accelerator is only a
memory proxy, not a compute coprocessor. When it detects a reference
to memory stored on a remote node, the accelerator would issue the
request through the network controller and track its progress,
delivering the data to the requesting thread when it was ready.

All global shared memory would be mapped into a segment of the local
address space on each node. The accelerator would do the
address-to-node translation. In addition, to support the issue and
complete operations for concurrency support, commands would be encoded
in the upper bits of the address: then each remote word would have multiple
addresses on the local node. Reading from the {\em issue} address would
immediately return a dummy value, but the accelerator would issue the
request over the network. Reading from the {\em complete} address
would block until the data was available.\todo{adjust with carl's suggestions}

@@

Implementing full-empty bit support efficiently on a platform not
designed for them is a challenge. It is possible to support full-empty
bit synchronization on arbitrary words by allocating additional
storage for the full-empty bits, and translating each atomic
full-empty operation into a sequence of memory operations that modify
the data and full-empty bit atomically, but this may have high
overhead. \todo{cite qthreads somehow}

@@

Synchronization operations can also be delegated to a memory manager
thread or accelerator. This may simplify the implementation of
synchronization operations: if only a single core (or single
accelerator) is accessing the data being synchronized, the use of
memory fences may be reduced or eliminated.

Note that we are not advocating using an FPGA as a memory controller
with attached DRAM and additional storage for full-empty bits. We
believe this unnecessarily complicates the design for such a
component. Our goal is to minimize the amount of hardware design our
proposal requires.
\todo{but are we eliminating the possibility of FE-only storage on the FPGA?}

@@

\begin{figure*}[ht]
%list walk ILP
\begin{minipage}[b]{0.3\linewidth}
\centering
\begin{alltt}{\small
  while (count-- > 0) \{
    list1 = list1->next;
    list2 = list2->next;
    \ldots
    list\(n\) = list\(n\)->next;
  \}
  }\centering{\bf (a)}
\end{alltt}
%\caption{Pseudocode for pointer chasing without coroutines.}
\label{fig:pointernocoro}
\end{minipage}
%listwalk green threads
\begin{minipage}[b]{0.35\linewidth}
\centering
\begin{alltt}{\small
  while (count-- > 0) \{
     prefetch(&(list1->next));
     switch();
     list1 = read(&(list1->next));
 \}
 }\centering{\bf (b)}
\end{alltt}
  %%n lists
  % while (count-- > 0) \{
  % prefetch(&(list1->next));
  %   prefetch(&(list2->next));
   %  \ldots
   %  prefetch(&(list\(n\)->next));
   %  switch();
   %  list1 = read(&(list1->next));
   %  list2 = read(&(list2->next));
   %  \ldots
   %  list\(n\) = read(&(list\(n\)->next));
%\caption{Pseudocode for pointer chasing using coroutines.}
\label{fig:pointercoro}
\end{minipage}
%cache pressure
\begin{minipage}[b]{0.32\linewidth}
\centering
\begin{alltt}{\small
  while (count-- > 0) \{
    prefetch(&(list1->next));
    switch();
    list1 = read(&(list1->next));
    for( i in 1 to num_local_updates ) \{
      local->data++;
      local = local->next;
   \} \}
   }\centering{\bf (c)}
\end{alltt}
%\caption{Pseudocode for pointer chasing with local updates.}
\label{fig:pointerupdate}
\end{minipage}
\vspace{10pt}
\caption{Pseudocode for: (a) pointer chasing without coroutines, (b) pointer chasing using coroutines, (c) pointer chasing with local updates. Coroutine code, as in (b), might also include ILP like what (a) depends on for memory concurrency, e.g., by doing multiple prefetches to independent lists, switch, then multiple reads.}
\label{fig:code}
\end{figure*}

@@

\begin{figure}[h]
	\begin{center}
		\includegraphics[width=0.47\textwidth]{figures/multi-listwalk-totalconc-edited.pdf}
	\end{center}
	\caption{Throughput of listwalk versus total number of
          concurrent references offered by the cores. Data points for
          running 1 core are shown as circles, and data points for
          running 6 cores are shown as diamonds. 
%Notice that the throughput for a single core levels out at 10
%concurrent references. For 4 to 6 cores, the throughput levels off at
%around 36 references, which seems to be the most memory concurrency a
%processor can handle.
        }
	\label{fig:listwalk-totalconc}
\end{figure}

@@

To simulate the performance of pointer chasing in a
multi-node system, we modified the scheduler of our coroutine library
to include a delay before a coroutine can be rescheduled, imitating
the network transit delay even though we are referencing local
memory. Recently published measurements of Infiniband latencies
\todo{cite} suggest we might expect 1.1 $\mu$s source-to-destination
latencies between two computers connected back-to-back; we assume a
switch transit delay of 400 ns \todo{how many hops? 1or3?} in each direction and thus estimate a 3
$\mu$s round-trip delay.

@@

There has been much work on user level threading; the First-Class User
Level Threading project \cite{ult} is one example. Our goal is
slightly different; we want many contexts to tolerate memory latency.

>>>>

%\documentclass{acm_proc_article-sp}
\documentclass[10pt,nocopyrightspace,preprint]{sigplanconf}
\usepackage{url}
\usepackage{fancyvrb}
\usepackage{alltt}
\usepackage{graphicx}
\usepackage{comment}
\usepackage[labelfont=bf,font=small,belowskip=-3pt,aboveskip=-5pt]{caption}
\usepackage[compact]{titlesec}
\usepackage{mdwlist}

\begin{document}

%\titlespacing{\section}{0pt}{*0}{*0}
%\titlespacing{\subsection}{0pt}{*0}{*0}
%\titlespacing{\subsubsection}{0pt}{*0}{*0}

%\textheight 9in
%\textwidth 6.5in
%\columnsep 1pc
\renewcommand{\baselinestretch}{0.85}

@@

\authorinfo{Jacob Nelson, Brandon Myers, A.H. Hunter, Preston Briggs,
  Dan Grossman, Mark Oskin, Carl Ebeling, Simon Kahan, Luis
  Ceze}{University of Washington}{\{nelson, bdmyers, ahh, preston,
  djg, oskin, ebeling, skahan, luisceze\}@cs.washington.edu}

@@

Multithreading is a technique that has been used successfully to
implement efficient computations for these graphs. The Cray XMT is an
example of such an approach: it solves the memory latency problem
through concurrency rather than caching. Each XMT processor supports
128 hardware contexts and 1024 outstanding memory operations, and is
able to switch contexts every cycle \cite{tera, feo-xmt}. This ability
comes at a cost: the XMT is an expensive, non-commodity machine.

\begin{figure*}[htbp]
  \begin{center}
    \vspace{-0.25in}
    \includegraphics[width=0.7\textwidth]{figures/system-overview.pdf}
    \vspace{-0.1in}
	\end{center}
	\caption{Overview of our proposed system design. The base
          system is a cluster of commodity nodes and interconnect. We
          add a latency-tolerant runtime and a global memory manager
          to obtain efficient access to the global address space. Only
          the shaded components may require hardware design.}
	\label{fig:system-overview}
\end{figure*}

@@

\section{Implementation}
\label{sec:approach}
Our goal is to use multithreading to tolerate
the latency of random accesses to memory in a global address space
spread across multiple nodes.  Our implementation must provide three features:
\begin{description}
\item[Lightweight multithreading:] We expect to need hundreds of
  threads to tolerated a cluster's network latency, so our threading
  implementation must have low overhead.

\item[Concurrency in global memory references:] A thread making a
  long-latency global memory reference must be able to yield and allow
  other threads to make progress.

\item[Low overhead synchronization:] 
  We must support global synchronization between threads that allows unrelated threads to proceed unblocked.
\end{description}
We discuss our approach to solving each of these challenges in
turn. For each one, we discuss both our current single-node
implementation and ideas for a multi-node implementation.

@@

\subsubsection{Single-node implementation}
We use prefetch instructions for the issue operation, and regular
blocking reads and writes for the complete operation.  Since we expect
these global references to have low locality, we use the
``non-temporal'' form of prefetch to avoid disturbing the state of the
cache hierarchy \cite{intel:swdev}.

@@

We believe we will need hundreds of outstanding memory references per
processor to cover the latency of remote references in a multi-node
system. Unfortunately, commodity processors have much smaller limits
on memory concurrency: section~\ref{subsubsec:evalsinglebase} finds a
limit of 36 concurrent loads. We will have to manage memory
concurrency on our own to avoid this limit. We describe approaches in
both software and hardware.

@@

%This approach has the benefit of enabling our entire system to be
%build using commodity parts, but its potential for performance is not
%clear. We have increased the overhead of a single memory operation:
%what in the single node case was a simple read, is now multiple
%queuing operations between threads and multiple reads and writes over
%the processor's link to the network interface.

Another approach moves the global memory management to a hardware
device. We can add a coprocessor FPGA in a processor socket, listening
to coherence messages on the bus, similar to \cite{mogill}. Note that
this accelerator is only a memory proxy, not a compute
coprocessor. The accelerator wold map all global shared memory into a
segment of each node's local address space; when it detects a
reference to memory on a remote node, it would do the address-to-node
translation and issue the request through the network controller,
delivering the data to the requesting thread when ready. We would
encode commands in the upper bits of the address; a read from a
location's {\em issue} address start the remote request and return
immediately, and a read from a location's {\em complete} address would
block until the data was available.

%To support the issue and complete operations for concurrency support,
%commands would be encoded in the upper bits of the address: then each
%remote word would have multiple addresses on the local node. Reading
%from the {\em issue} address would immediately return a value
%indicating that the accelerator is issuing the request over the
%network. Reading from the {\em complete} address would block until the
%data was available.

@@

Implementing full-empty bit support efficiently on a platform not
designed for them is a challenge. Previous work \cite{qthreads} has
enabled support full-empty bit synchronization on arbitrary words by
allocating additional storage for the full-empty bits and implementing
atomic full-empty operations as library routines.

@@

Synchronization operations can also be delegated to a memory manager
thread or accelerator. Sychronization on remote data would be
performed by the remote memory manager. This may simplify the
implementation: if only a single core (or single accelerator) is
accessing the data being synchronized, the use of memory fences may be
reduced or eliminated.

@@

\begin{figure*}[ht]
\vspace{-0.2in}
%list walk ILP
\begin{minipage}[b]{0.3\linewidth}{\small
\centering
\begin{alltt}{\scriptsize
  while (count-- > 0) \{
    list1 = list1->next;
    list2 = list2->next;
    \ldots
    list\(n\) = list\(n\)->next;
  \}
  }\centering{\bf (a)}
\end{alltt}
%\caption{Pseudocode for pointer chasing without coroutines.}
\label{fig:pointernocoro}
}\end{minipage}
%listwalk green threads
\begin{minipage}[b]{0.35\linewidth}{\small
\centering
\begin{alltt}{\scriptsize
  while (count-- > 0) \{
     prefetch(&(list1->next));
     switch();
     list1 = read(&(list1->next));
 \}
 }\centering{\bf (b)}
\end{alltt}
  %%n lists
  % while (count-- > 0) \{
  % prefetch(&(list1->next));
  %   prefetch(&(list2->next));
   %  \ldots
   %  prefetch(&(list\(n\)->next));
   %  switch();
   %  list1 = read(&(list1->next));
   %  list2 = read(&(list2->next));
   %  \ldots
   %  list\(n\) = read(&(list\(n\)->next));
%\caption{Pseudocode for pointer chasing using coroutines.}
\label{fig:pointercoro}
}\end{minipage}
%cache pressure
\begin{minipage}[b]{0.32\linewidth}{\small
\centering
\begin{alltt}{\scriptsize
  while (count-- > 0) \{
    prefetch(&(list1->next));
    switch();
    list1 = read(&(list1->next));
    for( i in 1 to num_local_updates ) \{
      local->data++;
      local = local->next;
   \} \}
   }\centering{\bf (c)}
\end{alltt}
%\caption{Pseudocode for pointer chasing with local updates.}
\label{fig:pointerupdate}
}\end{minipage}
%\vspace{5pt}
\caption{Pseudocode for: (a) pointer chasing without coroutines, (b) pointer chasing using coroutines, (c) pointer chasing with local updates. Coroutine code, as in (b), might also include ILP like what (a) depends on for memory concurrency, e.g., by doing multiple prefetches to independent lists, switch, then multiple reads.}
\label{fig:code}
\end{figure*}

@@

\begin{figure}[h]
  \vspace{-.1in}
	\begin{center}
		\includegraphics[width=0.47\textwidth]{figures/multi-listwalk-totalconc-edited.pdf}
	\end{center}
	\caption{Throughput of listwalk versus total number of
          concurrent references offered by the cores. Data points for
          running 1 core are shown as circles, and data points for
          running 6 cores are shown as diamonds. 
%Notice that the throughput for a single core levels out at 10
%concurrent references. For 4 to 6 cores, the throughput levels off at
%around 36 references, which seems to be the most memory concurrency a
%processor can handle.
        }
	\label{fig:listwalk-totalconc}
\end{figure}

@@

To simulate the performance of pointer chasing in a multi-node system,
we modified the scheduler of our coroutine library to include a delay
before a coroutine can be rescheduled, imitating the network transit
delay even though we are referencing local memory. We assume 1.1
$\mu$s interface and 400 ns switch latencies in each direction
\cite{infiniband, mellanox:site}, and thus estimate a 3 $\mu$s round-trip
delay.

@@

There has also been much work on user level threading; the First-Class
User Level Threading project \cite{ult} is one example. Our goal is
slightly different; we want many contexts to tolerate memory latency.
